---
title: "example_script_MKtest.Rmd"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(kableExtra)
```

# Load Janet's MK functions

Also loads the Biostrings package. 

```{r, warning=FALSE, message=FALSE}
source(here("scripts/MKfunctions.R"))
```


# Read in alignment, figure out sequence groupings

Read in alignment file, and fix the seqnames to remove descriptions.

This is an example alignment file I obtained from the MKT website. It contains four sequences of the ASIP gene: two seqs from Pongo pygmaeus (orangutan) and two from Trachypithecus cristatus (silvery lutung, an Old World monkey).


```{r}
alnFile <- here("test_data/MKTwebsiteExample/MKTwebsite_testAln.fa")
aln <- readBStringSet(alnFile)
aln
```
At first when we read in the alignment, sequence names also contain a description ("gi|909919..."): those are messy so I strip them off:

```{r}
names(aln) <- sapply( strsplit(names(aln), " "), "[[", 1)
aln
```

Figure out the names of sequences from each population. This step will be different for your alignment.

```{r}
pongo_names <- grep("pongo", names(aln), value=TRUE)
pongo_names
```

```{r}
trachy_names <- grep("trachy", names(aln), value=TRUE)
trachy_names
```


## Do the MK test

Do the MK test - simplest example:

```{r}
MKresults_websiteExample <- doMKtest(
    myAlnFile=alnFile, 
    pop1seqs=pongo_names, 
    pop2seqs=trachy_names)
```

The output of the doMKtest function (here, stored in the `MKresults_websiteExample` object) is a list, comprising two tables:      
- list element 1 ("summary")  contains the MK test results.   
- list element 2 ("positions")   contains a per-position analysis of the alignment, wihch helps understand what's going on with the alignment, and/or with troubleshooting.   


Here's the `summary` table, transposed to a single column for easier viewing. The results are at the bottom, with a record of the input dataset and parameters used at the top.

```{r}
MKresults_websiteExample[["summary"]] %>% 
    t() %>% 
    kable() %>% 
    kable_styling()
```


For p-values, to test for departures from neutrality, some people choose a chi-squared test  (e.g. the MKT website), other people prefer a Fishers exact test ("FET").  My script gives both p-values, so the user can choose which to report. I think FET is more conservative.

I also wrote a little function called `showContingencyTable` to pull out just the 2x2 contingency table showing counts from that summary table. Here's that table in the format used by the MKT website people:

```{r}
showContingencyTable(MKresults_websiteExample[["summary"]]) %>% 
    kable() %>% 
    kable_styling()
```

And here it is transposed (I prefer this:)

```{r}
showContingencyTable(MKresults_websiteExample[["summary"]]) %>% 
    t() %>% 
    kable() %>% 
    kable_styling()
```




Here's the first 6 rows of the `positions` output table, transposed to a single column for easier viewing

```{r}
MKresults_websiteExample[["positions"]] %>% 
    as_tibble() %>% 
    head() %>% 
    kable() %>% 
    kable_styling()
```

The positions table is mostly for troubleshooting code - it's more detail than most users need, although if you want to dig into which codons contain changes, this would be one way to do it.

## Saving results to Excel

Now, we do the same MK test but save the results in an Excel file  (we add the `writeMKoutput`, `outDir` and `outfileStem` options).  If we also use aliases, as below, it'll help us interpret the column names in the output.

As well as the list in R, we now also have an Excel output file called `MKTwebsite_testAln.MK.xlsx` (in the `test_data/MKTwebsiteExample` folder) There are two tabs, representing the two data.frames in the list output

Note that here I also use the `quiet=TRUE` option to suppress the status messages that the `doMKtest` function outputs while it's running

```{r}
MKresults_websiteExample <- doMKtest(
    myAlnFile=alnFile, 
    writeMKoutput=TRUE, 
    outDir = here("test_data/MKTwebsiteExample"), 
    outfileStem = "MKresults",
    pop1seqs=pongo_names, pop1alias = "pongo",
    pop2seqs=trachy_names, pop2alias = "trachy",
    quiet=TRUE)
```

## More options for the doMKtest function

There are some useful options in the `doMKtest` function.

### Filtering variants by frequency

We can filter rare variants using the `filterRareAlleles`/`alleleFreqThreshold` options:

(the example alignment only has two individuals for each population, so the filtering makes no sense for this example, and does not change the result)

```{r}
MKresults_websiteExample_removeRareVariants <- doMKtest(
    myAlnFile=alnFile, 
    pop1seqs=pongo_names, pop1alias = "pongo",
    pop2seqs=trachy_names, pop2alias = "trachy",
    filterRareAlleles=TRUE, alleleFreqThreshold=0.05,
    quiet=TRUE)
```

### Looking only at subregions of the alignment

We can look at only a certain region of the alignment, using the `regionStartAA`/`regionEndAA` options (coordinates are in amino acids, with respect to the alignment:

```{r}
MKresults_websiteExample_region1 <- doMKtest(
    myAlnFile=alnFile, 
    pop1seqs=pongo_names, pop1alias = "pongo",
    pop2seqs=trachy_names, pop2alias = "trachy",
    regionStartAA=1, regionEndAA=21,
    quiet=TRUE)
```

# Polarized MK tests

We can do polarized MK tests, if we supply an outgroup.

First, we read a different example alignment, of the Nicknack gene (CG17802) from *D. melanogaster* and *D. simulans*, with *D. yakuba* as an outgroup. Results published [here](https://elifesciences.org/articles/63368).

```{r}
nnk_alnFile <- here("test_data/Nnk_Kasinathan_2020/802.plusOutgroup_aln1_NT.edit.noStops.degapcodon.fas")
nnk_aln <- readBStringSet(nnk_alnFile)
nnk_aln
```

Then we get the names of sequences in each population:
```{r}
nnk_Dmel_names <- names(nnk_aln)[3:length(nnk_aln)]
nnk_Dsim_name <- names(nnk_aln)[2]
nnk_Dyak_names <- names(nnk_aln)[1]
```

Now we can use the `doMKtest` function, adding the `polarize=TRUE` option, and supplying outgroup seqnames via the `outgroupSeqs` option:

```{r}
nnk_MKresults_polarized <- doMKtest(
    myAlnFile=nnk_alnFile, 
    pop1seqs=nnk_Dmel_names, pop1alias = "Dmel",
    pop2seqs=nnk_Dsim_name, pop2alias = "Dsim",
    polarize=TRUE,
    outgroupSeqs=nnk_Dyak_names,
    quiet=TRUE )
```

# Combining results from multiple tests

If we've performed more than one MK test, we can combine the results using the `combineMKresults` function. The R object that's returned will be a single data.frame, combining the results. The Excel file will have the same table summarizing all the MK results, and one additional tab for each MK alignment showing each positions table.

```{r}
MKresults_combined <- combineMKresults(list(results1=MKresults_websiteExample,
                                            results2=nnk_MKresults_polarized),
                                       outFile="combinedMKresults.xlsx",
                                       outDir=here("test_data/test_output"))
```

# Plotting changes

I also made a plotting function called `plotMKpositions` to show where the fixed and polymorphic changes are:

Plot for the simple primate example:

```{r, fig.height=2, fig.width=7}
temp <- plotMKpositions(MKresults_websiteExample[["positions"]], 
                title="MKresults websiteExample", 
                pop1alias="pongo", pop2alias="trachy", 
                setNumPlots=FALSE)
```

Plot for polarized data, Nicknack example:

```{r, fig.height=5, fig.width=7}
temp <- plotMKpositions(nnk_MKresults_polarized[["positions"]], 
                        title="Nnk MK test", 
                        plotPolarizedPop1=TRUE, plotPolarizedPop2=TRUE, 
                        pop1alias="Dmel", pop2alias="Dsim")
```


# Finished

Show sessionInfo, to record package versions in case we need to do any troubleshooting.


```{r}
sessionInfo()
```

